Taking uncertainty in word embedding bias estimation seriously: a  bayesian

### 2023-05-17 paper/paperDraft.Rmd
revised first two figures

In what follows, we focus on  two popular  measures of bias applicable to many existing word embeddings ( \emph{GoogleNews},\footnote{GoogleNews-vectors-negative300, available at  \url{https://github.com/mmihaltz/word2vec-GoogleNews-vectors}.} \emph{Glove}\footnote{Available at \url{https://nlp.stanford.edu/projects/glove/}.} and \emph{Reddit Corpus}\footnote{The Reddit-L2 corpus, available at  \url{http://cl.haifa.ac.il/projects/L2/}.}): \emph{Word Embedding Association Test}  (WEAT) [@Caliskan2017semanticsBiases],  and \emph{Mean Average Cosine Distance} (MAC)  [@Manzini2019blackToCriminal].  We first explain how these measures are supposed to work. Then we argue that they are problematic for various reasons, a key one being that by pre-averaging data they manufacture false confidence, which we illustrate in terms of simulations showing that the measures often suggest the existence of bias even if by design it is non-existent in a simulated dataset.

We  propose to replace them with a Bayesian data analysis (hierarchical models), which not only provides more modest and realistic assessment of the uncertainty involved, but also allows for inspection at various levels of granularity. Once we introduce the method, we apply it to multiple word embeddings and results of supposed debiasing, putting forward some general observations that are not exactly in line with the usual picture painted in terms of WEAT or MAC (and the problem generalizes to any approach that focuses on chasing a single numeric metric): the word list sizes and sample sizes used in the studies are usually small,\footnote{Depending on a list for [@Caliskan2017semanticsBiases] the range for protected words is between 13 and 100, and for attributes between 16 and 25; for [@Manzini2019blackToCriminal] the range for protected words is between 14 and 18, and for attributes between 11 and 25}  posterior density intervals are fairly wide,   often the differences between associated, different and neutral human predicates, are not very impressive. Also, a preliminary inspection suggests that  the desirability of changes obtained by the usual debiasing methods is debatable. The tools that we have proposed, however, allow for a more fine-grained and multi-level evaluation of bias and debiasing in language models without abandoning modesty about the uncertainties involved.

@@

A measure of bias in word embeddings which does not employ similarities to  pre-calculated protected-class-level directions, the Word Embedding Association Test (WEAT), has been proposed in  [@Caliskan2017semanticsBiases]. The idea here is that the   bias between two sets of target words, $X$ and $Y$ (we call them protected words), should be quantified in terms of the cosine similarity  between the protected words and attribute words coming from  two sets of stereotype attribute words, $A$ and $B$ (we will call them attributes). For instance, $X$ might be a set of male names, $Y$ a set of female names, $A$ might contain stereotypically male-related, and $B$ stereotypically female-related career words. The association difference for a particular word $t$ (belonging to either $X$ or $Y$) is:

@@

\noindent The authors use it as a test statistic in some tests, and the assumption that $X$ and $Y$ are of the same size is necessary for the statistic to make sense. In the final measure (of effect size), WEAT, means are taken, which makes it insensitive to size differences between $X$ and $Y$:
\begin{align}
\mathsf{weat}(A,B) & = \frac{
\mu(\{\mathsf{s}(x,A,B)\}_{x\in X}) -\mu(\{\mathsf{s}(y,A,B)\}_{y\in Y}) 
}{
\sigma(\{\mathsf{s}(w,A,B)\}_{w\in X\cup Y})
}
\end{align}

 WEAT is inspired by  the Implicit Association Test (IAT) [@Nosek2002harvesting] used in psychology, and it uses almost the same word sets, allowing for a \emph{prima facie} sensible comparison with bias in humans. In  [@Caliskan2017semanticsBiases] the authors argue that significant biases---thus measured--- similar to the ones discovered by IAT can be discovered in word embeddings. In [@Lauscher2019multidimensional] the methodology is  extended to a multilingual and cross-lingual setting, arguing that using Euclidean distance instead of cosine similarity does not make much difference, while the bias effects vary greatly across embedding models (interestingly, with social media-text trained embeddings being less biased than those based on Wikipedia).
 A similar methodology is employed in [@Garg2018years]. The authors employ word embeddings trained on corpora from different decades to study the shifts in various biases.^[Strictly speaking, these authors use Euclidean distances and their differences, but the way they take averages and averages thereof is analogous, and so what we will have to say about pre-averaging leading to false confidence applies to this methodology as well.]

\begin{figure}[H]
```{r WEATexample,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "55%", warning = FALSE, message = FALSE}
words <- c("man", "he", "boss", "architect","nurse","receptionist","woman","she")
x <- c(.1,.1,.5,.5,.7,.7,1.2,1.2)
y <- c(.4,.3,.3,.2,.5,.4,.4,.3)
weat <- data.frame(words,x,y)

ggplot(weat, aes(x=x,y=y, label = words))+geom_point()+geom_text(check_overlap = TRUE,
                                                                 nudge_y = 0.018, size = 7 )+
    geom_segment(data=weat, mapping=aes(x=.1, y=.3, xend=.5, yend=.2),  linetype =2, arrow=arrow(), size=.05, color="blue") +
  geom_segment(data=weat, mapping=aes(x=.1, y=.3, xend=.5, yend=.3), linetype =2, arrow=arrow(), size=.05, color="blue")+
  geom_segment(data=weat, mapping=aes(x=.1, y=.3, xend=.7, yend=.4), linetype =2, arrow=arrow(), size=.05, color="orangered")+
  geom_segment(data=weat, mapping=aes(x=.1, y=.3, xend=.7, yend=.5), linetype =2, arrow=arrow(), size=.05, color="orangered")+
    annotate("text", x = c(.32,.32,.32,.32), y = c(.23,.29,.35,.39), 
             label = c(".7", ".6",".1", ".2") , color="black", 
             size=7,  fontface="bold")+
    annotate("text", x = c(.1,.5,.7,1.2), y = c(.55,.55,.55,.55), 
           label = c("X", "A","B", "Y") , color=c("black", "blue","orangered","black"), 
           size=10,  fontface="bold")+xlim(0,1.3)+
  theme_void(base_size = 25)
```
\caption{Example WEAT calculations: $s_1 = s(he,A,B) =  \frac{.6+.7}{2}  - \frac{.2+.1}{2} = .65-.15= .5$,  $s_2 = s(man,A,B) = .3$,  $s_3 = s(woman,A,B) = -.6, s_4 = s(she, A, B) = -.3$, $\mathsf{WEAT}(A,B)  = \frac{\nicefrac{(s_1+s_2)}{2} - \nicefrac{(s_3+s_4)}{2}}{sd(\{s_1,s_2,s_3,s_4\})} \approx 1.93$.}
\label{eq:WEATexample}
\end{figure}

- $s_1 = s(he,A,B) =  \frac{.6+.7}{2}  - \frac{.2+.1}{2} = .65-.15= .5$

- $s_2 = s(man,A,B) = .3$, \linebreak  $s_3 = s(woman,A,B) = -.6, s_4 = s(she, A, B) = -.3$

\vspace{-4mm}

\normalsize 
\begin{align*}
\mathsf{WEAT}(A,B) & = \frac{\nicefrac{(s_1+s_2)}{2} - \nicefrac{(s_3+s_4)}{2}}{sd(\{s_1,s_2,s_3,s_4\})} \approx 1.93
\end{align*}

WEAT, however, has been developed to investigate biases corresponding to a pair of supposedly opposing stereotypes, and so the question arises as to how to generalize the measure to contexts in which biases with respect to more than two stereotypical groups are to be measured. Such a generalization can be found in [@Manzini2019blackToCriminal]. The authors introduce Mean Average Cosine distance (MAC) as a measure of bias. Let $T = \{t_1, \dots, t_k\}$ be a set of protected words, and let each $A_j\in \mathcal{A}$ be a set of attributes stereotypically associated with a protected word where $\mathcal{A}$.
For instance, when biases related to religion are to be investigated, they use a dataset of the  format illustrated in Table \ref{tab:religionOriginal}. The measure is defined as follows:

@@

```
\caption{Example of MAC calculations.  $s_1  = s(muslim,A_1) = \frac{cos(muslim,dirty)+cos(muslim,terrorist)}{2}$, 
$s_2  = s(muslim,A_2) = \frac{cos(muslim,familiar)+cos(muslim,conservative)}{2}$, \dots. Then  $\mathsf{MAC}(T,A)  = \mathsf{mean}(\{s_i \vert i \in 1, \dots, k\})$.}
\label{fig:MACexample}
\end{figure}

@@

Here is an example of results of debiasing that can be found in [@Manzini2019blackToCriminal] (Table \ref{tab:religionOriginal2}). It presents the results of applying the procedure with respect to the Religion dataset. MAC scores are given for the original datasets, and two resulting from two debiasing methods applied to the Reddit embedding. The scores for debiasing come with p-values for the comparison to the original MAC.

@@

\begin{tabular}[t]{lllr}
\toprule
Religion Debiasing & MAC & a p-value \\
\midrule
\cellcolor{gray!15}{Biased} & \cellcolor{gray!15}{0.859} & \cellcolor{gray!15}{N/A} \\
Hard Debiased & 0.934 & 3.006e-07\\
\cellcolor{gray!15}{Soft Debiased ($\lambda$ = 0.2)} & \cellcolor{gray!15}{0.894} & \cellcolor{gray!15}{0.007} \\
\bottomrule
\end{tabular}

\caption{The associated mean average cosine similarity
(MAC) and p-values for debiasing methods for religious bias.}
\label{tab:religionOriginal2}
\normalsize 
\end{table}

The first conceptual question we should ask is whether   the initial MAC values lower than 1 indeed are  indicative of the presence of bias? Of course, thinking abstractly, 1 is the ideal distance for unrelated words. But clearly, there is variation in distances, which might lead to non-biased lists also having MAC smaller than 1. How much smaller?  What may attract attention is the fact that the value of cosine distance in "Biased" category is already quite high (i.e. close to 1) even before debiasing. High cosine distance indicates low cosine similarity between values. One could think that average cosine similarity equal to approximately 0.141 is not large enough to consider it as bias. The authors, though, still aim to mitigate such "bias" to make the distance even larger. The question is, on what basis is this small similarity still considered as a proof of the presence of bias, and whether these small changes are meaningful.

The problem is that the original paper did not employ any control group of neutral attributes for comparison to obtain a more realistic gauge on how to understand MAC values. Later on, in our approach, we introduce such control word lists. One of them is a list of words we intuitively considered neutral (see the Appendix for word lists).

Moreover, it might be the case that words that have to do with human activities in general, even if unbiased, are systematically closer to the protected words than merely neutral words. This, again, casts doubt on whether comparing MAC to the abstractly ideal value of 1 is a methodologically sound idea. For this reason we also use a second list with intuitively non-stereotypical human attributes (again, see the Appendix for a word list).

Another important observation is that  MAC calculations  do not distinguish whether a given  attribute is associated with a given protected word, simply averaging across all such groups.  Let us use the  case of religion-related stereotypes to illustrate. The full lists from [@Manzini2019blackToCriminal] can be found in Appendix \ref{appendix:manzini_word_lists}. In the original paper,  words from all three religions were compared against all of the stereotypes. No distinction between cases in which the stereotype is associated with a given religion, as opposed to the situation in which it is associated with another one, is made. For example, the protected word \emph{jew} is supposed to be stereotypically  connected with  the attribute \emph{greedy}, while  from the protected word \emph{quran} the  attribute \emph{greedy} comes from a different stereotype, and yet the distances between these pairs contribute equally to the final \textsf{MAC} score. This is problematic, as  not all of the stereotypical words have to be considered as harmful for all of the religions. To avoid the masking effect, one should  pay attention to how protected words and attributes are paired by stereotypes.

In Figure  \ref{fig:empirical} we look at the empirical distributions, while paying attention to such divisions. The horizontal lines represent the values of 1-\textsf{MAC} (that is, we now talk in terms of cosine similarity rather than cosine distance; this is a trivial linear transformation) that the authors considered indicative of bias for stereotypes corresponding to  given  word lists. For instance, in religion, MAC was .859 and considered a sign of bias, so we plot $0\pm (1-.859)\approx .14$ lines around similarity = 0 (that is, distance = 1). Notice that most distributions are quite wide, and the proportions of even neutral or human neutral words with similarities higher than the value of 1-\textsf{MAC} deserving debiasing according to the authors is quite high.

@@

The t-tests they employ are run on average cosines used to calculate MAC.

@@

In the calculations of WEAT means are taken twice. The $s$-values themselves are means, and then means of $s$-values are compared between groups. The test statistic used itself is of this sort as well---while it compares sums, these are sums of sets of the same size, so they are really linear transformations of means of means. Statistical troubles start when we run statistical tests on sets of means, for at least two reasons  One, by pre-averaging data we throw away information about sample sizes. For the former point, think about proportions: 10 out of 20 and 2 out of 4 give the same mean, but you would obtain more information by making the former observation rather than by making the latter.  And especially in this context, in which the word lists are not huge, sample sizes should matter. Two, when we pre-average, we remove variation, and so pre-averaging  tend to manufacture false confidence. Group means display less variation than the raw data points, and the standard deviation of a set of means of sets of means is bound to be lower than the original standard deviation in the row data. Now, if you calculate your effect size by dividing by the pre-averaged standard deviation, your quite likely to get something that looks like a strong effect size, but the results of your calculations might not track anything interesting.

@@

Where do we stand now? We have seen  that seemingly high effect sizes  might arise even if the underlying processes actually have the same mean.  The uncertainty resulting from including the raw data point variance into considerations is more extensive than the one suggested by the low p-values obtained from taking means or means of means as data points. In this section, we discussed the performance of the WEAT measure, but since the [@Manzini2019blackToCriminal] one is a generalization thereof, including the method of running statistical tests on pre-averaged data, our remarks, \emph{mutatis nutandis}, apply. Moreover, a general problem is that studies such as [@Caliskan2017semanticsBiases] or [@Manzini2019blackToCriminal]  do not use  control groups, and this makes the interpretation of the analysis additionally difficult.

@@

After introducing the method, we apply it to multiple word embeddings and results of supposed debiasing, putting forward some general observations that are not exactly in line with the usual picture painted in terms of WEAT or MAC (and the problem generalizes to any approach that focuses on chasing a single numeric metric): the word list sizes and sample sizes used in the studies are usually small,^[Depending on a list for [@Caliskan2017semanticsBiases] the range for protected words is between 13 and 100, and for attributes between 16 and 25; for [@Manzini2019blackToCriminal] the range for protected words is between 14 and 18, and for attributes between 11 and 25.]  posterior density intervals are fairly wide,   often the differences between associated, different and neutral human predicates, are not very impressive. Also, a preliminary inspection suggests that  the desirability of changes obtained by the usual debiasing methods is debatable. The tools that we propose, however, allow for a more fine-grained and multi-level evaluation of bias and debiasing in language models without losing modesty about the uncertainties involved.

>>>>

In what follows, we focus on  two popular  measures of bias applicable to many existing word embeddings ( \emph{GoogleNews},\footnote{GoogleNews-vectors-negative300, available at  \url{https://github.com/mmihaltz/word2vec-GoogleNews-vectors}.} \emph{Glove}\footnote{Available at \url{https://nlp.stanford.edu/projects/glove/}.} and \emph{Reddit Corpus}\footnote{The Reddit-L2 corpus, available at  \url{http://cl.haifa.ac.il/projects/L2/}.}): \emph{Word Embedding Association Test}  (\textsf{WEAT}) [@Caliskan2017semanticsBiases],  and \emph{Mean Average Cosine Distance} (\textsf{MAC})  [@Manzini2019blackToCriminal].  We first explain how these measures are supposed to work. Then we argue that they are problematic for various reasons, a key one being that by pre-averaging data they manufacture false confidence, which we illustrate in terms of simulations showing that the measures often suggest the existence of bias even if by design it is non-existent in a simulated dataset.

We  propose to replace them with a Bayesian data analysis (hierarchical models), which not only provides more modest and realistic assessment of the uncertainty involved, but also allows for inspection at various levels of granularity. Once we introduce the method, we apply it to multiple word embeddings and results of supposed debiasing, putting forward some general observations that are not exactly in line with the usual picture painted in terms of \textsf{WEAT} or \textsf{MAC} (and the problem generalizes to any approach that focuses on chasing a single numeric metric): the word list sizes and sample sizes used in the studies are usually small,^[Depending on a list for [@Caliskan2017semanticsBiases] the range for protected words is between 13 and 100, and for attributes between 16 and 25; for [@Manzini2019blackToCriminal] the range for protected words is between 14 and 18, and for attributes between 11 and 25.]  posterior density intervals are fairly wide,   often the differences between associated, different and neutral human predicates, are not very impressive. Also, a preliminary inspection suggests that  the desirability of changes obtained by the usual debiasing methods is debatable. The tools that we have proposed, however, allow for a more fine-grained and multi-level evaluation of bias and debiasing in language models without abandoning modesty about the uncertainties involved.

@@

A measure of bias in word embeddings which does not employ similarities to  pre-calculated protected-class-level directions, the Word Embedding Association Test (\textsf{WEAT}), has been proposed in  [@Caliskan2017semanticsBiases]. The idea here is that the   bias between two sets of target words, $X$ and $Y$ (we call them protected words), should be quantified in terms of the cosine similarity  between the protected words and attribute words coming from  two sets of stereotype attribute words, $A$ and $B$ (we will call them attributes). For instance, $X$ might be a set of male names, $Y$ a set of female names, $A$ might contain stereotypically male-related, and $B$ stereotypically female-related career words. The association difference for a particular word $t$ (belonging to either $X$ or $Y$) is:

@@

\noindent The authors use it as a test statistic in some tests, and the assumption that $X$ and $Y$ are of the same size is necessary for the statistic to make sense. In the final measure (of effect size), \textsf{WEAT}, means are taken, which makes it insensitive to size differences between $X$ and $Y$:
\begin{align}
\mathsf{weat}(A,B) & = \frac{
\mu(\{\mathsf{s}(x,A,B)\}_{x\in X}) -\mu(\{\mathsf{s}(y,A,B)\}_{y\in Y}) 
}{
\sigma(\{\mathsf{s}(w,A,B)\}_{w\in X\cup Y})
}
\end{align}

 \textsf{WEAT} is inspired by  the Implicit Association Test (IAT) [@Nosek2002harvesting] used in psychology, and it uses almost the same word sets, allowing for a \emph{prima facie} sensible comparison with bias in humans. In  [@Caliskan2017semanticsBiases] the authors argue that significant biases---thus measured--- similar to the ones discovered by IAT can be discovered in word embeddings. In [@Lauscher2019multidimensional] the methodology is  extended to a multilingual and cross-lingual setting, arguing that using Euclidean distance instead of cosine similarity does not make much difference, while the bias effects vary greatly across embedding models (interestingly, with social media-text trained embeddings being less biased than those based on Wikipedia).
 A similar methodology is employed in [@Garg2018years]. The authors employ word embeddings trained on corpora from different decades to study the shifts in various biases.^[Strictly speaking, these authors use Euclidean distances and their differences, but the way they take averages and averages thereof is analogous, and so what we will have to say about pre-averaging leading to false confidence applies to this methodology as well.]

\begin{figure}[H]
```{r WEATexample5,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "55%", warning = FALSE, message = FALSE}
words <- c("man", "he", "boss", "architect","nurse","receptionist","woman","she")
x <- c(.1,.1,.5,.5,.7,.7,1.2,1.2)
y <- c(.4,.3,.3,.2,.5,.4,.4,.3)
weat <- data.frame(words,x,y)

ggplot(weat, aes(x=x,y=y, label = words))+geom_point()+geom_text(check_overlap = TRUE,
                                                                 nudge_y = 0.021, size = 7 )+
    geom_segment(data=weat, mapping=aes(x=.1, y=.3, xend=.5, yend=.2),  linetype =2, arrow=arrow(), size=.05, color="blue") +
  geom_segment(data=weat, mapping=aes(x=.1, y=.3, xend=.5, yend=.3), linetype =2, arrow=arrow(), size=.05, color="blue")+
  geom_segment(data=weat, mapping=aes(x=.1, y=.3, xend=.7, yend=.4), linetype =2, arrow=arrow(), size=.05, color="orangered")+
  geom_segment(data=weat, mapping=aes(x=.1, y=.3, xend=.7, yend=.5), linetype =2, arrow=arrow(), size=.05, color="orangered")+
    annotate("text", x = c(.32,.32,.32,.32), y = c(.23,.29,.35,.39), 
             label = c(".7", ".6",".1", ".2") , color="black", 
             size=7,  fontface="bold")+
    annotate("text", x = c(.1,.5,.7,1.2), y = c(.55,.55,.55,.55), 
           label = c("X", "A","B", "Y") , color=c("black", "blue","orangered","black"), 
           size=10,  fontface="bold")+xlim(0,1.3)+
  theme_void(base_size = 25)
```
\caption{A simple example of a two-class set-up. Two groups of protected words, $X$ and $Y$ with two stereotypical attribute sets. An example of \textsf{WEAT} calculations follows.}
\label{eq:WEATexample}
\end{figure}

\noindent Here is an example of \textsf{WEAT} calculations for Figure \ref{eq:WEATexample}:
\begin{align*} s_1 & = s(he,A,B)  =  \nicefrac{(.6+.7)}{2}  - \nicefrac{(.2+.1)}{2} = .65-.15= .5 \\
s_2  & = s(man,A,B) = .3 \\
s_3  & = s(woman,A,B) = -.6\\
s_4 & = s(she, A, B) = -.3\\
\mathsf{WEAT}(A,B)  & = \frac{\nicefrac{(s_1+s_2)}{2} - \nicefrac{(s_3+s_4)}{2}}{sd(\{s_1,s_2,s_3,s_4\})} \approx 1.93
\end{align*}

\textsf{WEAT}, however, has been developed to investigate biases corresponding to a pair of supposedly opposing stereotypes, and so the question arises as to how to generalize the measure to contexts in which biases with respect to more than two stereotypical groups are to be measured. Such a generalization can be found in [@Manzini2019blackToCriminal]. The authors introduce Mean Average Cosine distance (\textsf{MAC}) as a measure of bias. Let $T = \{t_1, \dots, t_k\}$ be a set of protected words, and let each $A_j\in \mathcal{A}$ be a set of attributes stereotypically associated with a protected word where $\mathcal{A}$.
For instance, when biases related to religion are to be investigated, they use a dataset of the  format illustrated in Table \ref{tab:religionOriginal}. The measure is defined as follows:

@@

```
\caption{A small subset of the religion dataset.  To each protected word in $T$  there corresponds one class of stereotypical attributes typically associated with it (and other classes of stereotypical attributes associated with different protected words).}
\label{fig:MACexample}
\end{figure}

\noindent An example of \textsf{MAC} calculations for the situation depicted in Figure \ref{fig:MACexample} is as follows:
\begin{align*}
s_1   = s(muslim,A_1)  & = \frac{\mathsf{cos}(muslim,dirty)+\mathsf{cos}(muslim,terrorist)}{2} \\ 
s_2   = s(muslim,A_2)  & = \frac{\mathsf{cos}(muslim,familiar)+\mathsf{cos}(muslim,conservative)}{2}\\ & \vdots \\ 
\mathsf{MAC}(T,A)  & = \mathsf{mean}(\{s_i \vert i \in 1, \dots, k\})
\end{align*}

Notably, the inutitive distinction between different attribute sets plays no real role in the \textsf{MAC} calculations. Equally well one could calculate the mean distance of \emph{muslim} to all the predicates, mean distance of \emph{christian} to all the predicates, mean distance of \emph{jew} to all the predicates, and then to take the mean of these three means.

@@

Here is an example of results of debiasing that can be found in [@Manzini2019blackToCriminal] (Table \ref{tab:religionOriginal2}). It presents the results of applying the procedure with respect to the Religion dataset. \textsf{MAC} scores are given for the original datasets, and two resulting from two debiasing methods applied to the Reddit embedding. The scores for debiasing come with p-values for the comparison to the original \textsf{MAC}.

@@

\begin{tabular}[t]{lllr}
\toprule
Religion Debiasing & \textsf{MAC} & a p-value \\
\midrule
\cellcolor{gray!15}{Biased} & \cellcolor{gray!15}{0.859} & \cellcolor{gray!15}{N/A} \\
Hard Debiased & 0.934 & 3.006e-07\\
\cellcolor{gray!15}{Soft Debiased ($\lambda$ = 0.2)} & \cellcolor{gray!15}{0.894} & \cellcolor{gray!15}{0.007} \\
\bottomrule
\end{tabular}

\caption{The associated mean average cosine similarity
(\textsf{MAC}) and p-values for debiasing methods for religious bias.}
\label{tab:religionOriginal2}
\normalsize 
\end{table}

The first conceptual question we should ask is whether   the initial \textsf{MAC} values lower than 1 indeed are  indicative of the presence of bias? Of course, thinking abstractly, 1 is the ideal distance for unrelated words. But clearly, there is variation in distances, which might lead to non-biased lists also having \textsf{MAC} smaller than 1. How much smaller?  What may attract attention is the fact that the value of cosine distance in "Biased" category is already quite high (i.e. close to 1) even before debiasing. High cosine distance indicates low cosine similarity between values. One could think that average cosine similarity equal to approximately 0.141 is not large enough to consider it as bias. The authors, though, still aim to mitigate such "bias" to make the distance even larger. The question is, on what basis is this small similarity still considered as a proof of the presence of bias, and whether these small changes are meaningful.

The problem is that the original paper did not employ any control group of neutral attributes for comparison to obtain a more realistic gauge on how to understand \textsf{MAC} values. Later on, in our approach, we introduce such control word lists. One of them is a list of words we intuitively considered neutral (see the Appendix for word lists).

Moreover, it might be the case that words that have to do with human activities in general, even if unbiased, are systematically closer to the protected words than merely neutral words. This, again, casts doubt on whether comparing \textsf{MAC} to the abstractly ideal value of 1 is a methodologically sound idea. For this reason we also use a second list with intuitively non-stereotypical human attributes (again, see the Appendix for a word list).

Another important observation is that  \textsf{MAC} calculations  do not distinguish whether a given  attribute is associated with a given protected word, simply averaging across all such groups.  Let us use the  case of religion-related stereotypes to illustrate. The full lists from [@Manzini2019blackToCriminal] can be found in Appendix \ref{appendix:manzini_word_lists}. In the original paper,  words from all three religions were compared against all of the stereotypes. No distinction between cases in which the stereotype is associated with a given religion, as opposed to the situation in which it is associated with another one, is made. For example, the protected word \emph{jew} is supposed to be stereotypically  connected with  the attribute \emph{greedy}, while  from the protected word \emph{quran} the  attribute \emph{greedy} comes from a different stereotype, and yet the distances between these pairs contribute equally to the final \textsf{MAC} score. This is problematic, as  not all of the stereotypical words have to be considered as harmful for all of the religions. To avoid the masking effect, one should  pay attention to how protected words and attributes are paired by stereotypes.

In Figure  \ref{fig:empirical} we look at the empirical distributions, while paying attention to such divisions. The horizontal lines represent the values of 1-\textsf{MAC} (that is, we now talk in terms of cosine similarity rather than cosine distance; this is a trivial linear transformation) that the authors considered indicative of bias for stereotypes corresponding to  given  word lists. For instance, in religion, \textsf{MAC} was .859 and considered a sign of bias, so we plot $0\pm (1-.859)\approx .14$ lines around similarity = 0 (that is, distance = 1). Notice that most distributions are quite wide, and the proportions of even neutral or human neutral words with similarities higher than the value of 1-\textsf{MAC} deserving debiasing according to the authors is quite high.

@@

The t-tests they employ are run on average cosines used to calculate \textsf{MAC}.

@@

In the calculations of \textsf{WEAT} means are taken twice. The $s$-values themselves are means, and then means of $s$-values are compared between groups. The test statistic used itself is of this sort as well---while it compares sums, these are sums of sets of the same size, so they are really linear transformations of means of means. Statistical troubles start when we run statistical tests on sets of means, for at least two reasons  One, by pre-averaging data we throw away information about sample sizes. For the former point, think about proportions: 10 out of 20 and 2 out of 4 give the same mean, but you would obtain more information by making the former observation rather than by making the latter.  And especially in this context, in which the word lists are not huge, sample sizes should matter. Two, when we pre-average, we remove variation, and so pre-averaging  tend to manufacture false confidence. Group means display less variation than the raw data points, and the standard deviation of a set of means of sets of means is bound to be lower than the original standard deviation in the row data. Now, if you calculate your effect size by dividing by the pre-averaged standard deviation, your quite likely to get something that looks like a strong effect size, but the results of your calculations might not track anything interesting.

@@

Where do we stand now? We have seen  that seemingly high effect sizes  might arise even if the underlying processes actually have the same mean.  The uncertainty resulting from including the raw data point variance into considerations is more extensive than the one suggested by the low p-values obtained from taking means or means of means as data points. In this section, we discussed the performance of the \textsf{WEAT} measure, but since the [@Manzini2019blackToCriminal] one is a generalization thereof, including the method of running statistical tests on pre-averaged data, our remarks, \emph{mutatis nutandis}, apply. Moreover, a general problem is that studies such as [@Caliskan2017semanticsBiases] or [@Manzini2019blackToCriminal]  do not use  control groups, and this makes the interpretation of the analysis additionally difficult.

@@

After introducing the method, we apply it to multiple word embeddings and results of supposed debiasing, putting forward some general observations that are not exactly in line with the usual picture painted in terms of \textsf{WEAT} or \textsf{MAC} (and the problem generalizes to any approach that focuses on chasing a single numeric metric): the word list sizes and sample sizes used in the studies are usually small.  posterior density intervals are fairly wide,   often the differences between associated, different and neutral human predicates, are not very impressive. Also, a preliminary inspection suggests that  the desirability of changes obtained by the usual debiasing methods is debatable. The tools that we propose, however, allow for a more fine-grained and multi-level evaluation of bias and debiasing in language models without losing modesty about the uncertainties involved.
